---
title: "Exam 2016"
author: "Ninette Lotter"
date: "31 August 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

**Alternative hypothesis:** Body temperature and heart rate is different between men and women

**Null hypothesis:** There is no difference in body temperature or heart rate between men and women.

```{r question_1, warning=FALSE}
# Load packages
library(tidyr)
library(dplyr)
library(ggplot2)
# Import data
data1 <- readr::read_csv('https://dl.dropboxusercontent.com/u/11805474/painblogr/biostats/exam-data-2016/question1.csv')
# Summary statistics
summary(data1)
head(data1)
# TIdy data
gathered1 <- gather(data1, key = Sex, value = Heart_rate, male:female, na.rm = T)
gathered1
summary(gathered1)
# Perform statistical test
t.test(body_temperature ~ Sex, data = gathered1, paired = T)
t.test(Heart_rate ~ Sex, data = gathered1, paired = T)
```

**Type of data:** Paired continuous data.

**Test Assumptions:**

* Data are normally distributed.
* Equal variance across groups.
* The errors are independent.
* Matching is effective.

**Statistical test:** t-test

**Test statistic:**

_Heart rate ~ Sex_
t = 0.577
p =0.566
95% CI = -1.93 to 3.501

_Body temperature ~ Sex_
t = 11.536
p < 0.001

**Conclusion:** There is no significant difference in heart rate (t = 0.577, p = 0.566) between men and women. Body temperature is significantly different (t = 11.536, p < 0.001) between men and women (95% CI does not include zero, therefore we can accept the null hypothesis regarding body temperature).

## Question 2

**Alternative hypothesis:** Handedness is associated with ataxic walking in intoxicated individuals.

**Null hypothesis:** There is no association between handedness and ataxic walking in intoxicated individuals.

```{r question_2, warning=FALSE}
# load packages
library(tidyr)
library(dplyr)
# Import data
data2 <- readr::read_csv('https://dl.dropboxusercontent.com/u/11805474/painblogr/biostats/exam-data-2016/question2.csv')
# Summary statistics
summary(data2)
head(data2)
# Select a subset of columns
select <- select(data2, handedness, final_position)
# Cross tabulate selected columns
tab <- xtabs(~ handedness + final_position, data = select)
tab
# Plot data
barplot(tab, beside = TRUE, legend.text = TRUE)
# Statistical test
mcnemar.test(tab)
```

**Type of data:** Paired ordinal data.

**Test Assumptions:**

* The two groups are mutually independent.
* Each x and y can be assigned to two possible categories.
* Random sampling.


**Statistical test:** McNemar Test

**Test statistic:**

Chi-squared = 1.0159
Degrees of freedom = 1
Sample size = 151
p = 0.3135

**Conclusion:** There is no significant association (Chi-squared = 1.0159, p = 0.3135) between handedness and ataxic walking in intoxicated individuals.

## Question 3

**Alternative hypothesis:** Running time is associated with calories consumed during a run.

**Null hypothesis:** There is no association between running time and calorie consumption.

```{r question_3, warning=FALSE}
# Load packages
library(readr)
library(tidyr)
# Import dataset
data3 <- readr::read_csv('https://dl.dropboxusercontent.com/u/11805474/painblogr/biostats/exam-data-2016/question3.csv')
# Summary statistics
summary(data3)
head(data3)
# Exploratory plot (modified to publishable format)
plot(data3$time, data3$calories, xlab = 'Running time (s)', ylab = 'Calories consumed (cal)', main = 'Correlation between running time and calories consumed during run.')
# Abline of linear regression
abline(lm(calories ~ time, data = data3))

# Test for correlation 
cor.test(data3$time, data3$calories, method = 'pearson')
# Linear regression analysis
model <- lm(calories ~ time, data = data3)
# Print summary of linear regression statistics
summary(model)
# Diagnostic plots to confirm linear regression
plot(x = model$fitted, y = model$residuals)
abline(h = 0)
qqnorm(model$residuals)
qqline(model$residuals)
```

**Type of data:** Unpaired continuous data.

**Assumptions for Pearson's correlation:** 

* There is a linear relationship between the variables.
* There are no outliers
* Both variables are approximately normally distributed.

**Assumptions for linear regression:**

* There is a linear trend between x and y.
* The observations in the sample are independent.
* x is measured without error.
* The residuals are normally distributed.
* The residuals have the same variance for all fitted values of y (homoskedastic).

**Statistical test:** Pearson's correlation and linear regression model.

**Results:**

_Correlation analysis_
Pearson's t = -3.77
Degrees of freedom = 17
p = 0.0015
95% Confidence Interval = -0.86 to -0.32

_Linear regression_
F-statistic = 14.22
Degrees of freedom = 17
p = 0.0015

The graph fitted for homoskedacity shows no pattern, therefore the residuals have the same variance for all fitted values of y. 
The quantile-quantile plot showed a Gaussian distribution of the residuals. 

**Conclusion:** There is a significant inverse relationship (t = -3.77, F = 14.22, p = 0.0015) between running time and the calories consumed during a run.

## Question 4

```{r question_4, warning=FALSE, dev='pdf', fig.path='./figures/'}
library(readr)
library(tidyr)
foo <- rnorm(10000, mean = 60, sd = 3) # final mark
bar <- rnorm(10000, mean = 68, sd = 5) # project mark
baz <- rep(seq(from = 1997, to = 2006), each = 1) # years

year <- sample(baz, 150, replace = TRUE,
               prob = c(0.05, 0.05, 0.08, 0.08, 
                          0.1, 0.13, 0.14, 0.13, 0.12, 0.12))
project_mark <- sample(bar, 150, replace = TRUE)
final_mark <- sample(foo, 150, replace = TRUE)

plot_data <- data_frame(year = year,
                        project_mark = round(project_mark, 1),
                        final_mark = round(final_mark, 1)) %>%
    arrange(year)
plot(plot_data$project_mark, plot_data$final_mark, xlab = 'Project mark (%)', ylab = 'Final mark (%)', main = 'Relationship between project mark and final mark.', ylim = c(50, 80), xlim = c(50, 80))
# Axis scale was set from 50 to 80 because the distribution and trend of the marks is easier to see.
```

